---
title: "Cannibalisation in the software product lines <br> RDD used for Windows OS"
output:
  github_document:
    html_preview: false
warnings: no
---
<b>Author:</b> Przemyslaw Niedziela (przemyslaw.niedziela98@gmail.com) <br> 
<b>Date:</b> May 2024 <br>
<br> <br> 
<b> TL;DR </b> <br>
This analysis aims to estimate the cannibalization effect following the launch of Windows 10 on Windows 8.1 and Windows 7, utilizing a quasi-RDD approach. This method seeks to estimate discontinuity and kinks among specific Windows OS. The data is limited to Europe, with the unit of analysis being households. <br>  
Data is sourced from two main providers: <br> <br> 
(1) Eurostat - two datasets: <br>
  -  [Households - availability of computers](https://ec.europa.eu/eurostat/web/products-datasets/-/isoc_ci_cm_h) <br> 
  -  [Number of households by household composition](https://data.europa.eu/data/datasets/jfpeap7gs0z4sxsrztrkig) <br> 
(2) StatCounter - two datasets: <br>
- [Desktop Windows Version Market Share Europe](https://gs.statcounter.com/os-version-market-share/windows/desktop/europe#monthly-200902-202405) <br> 
- [Operating System Market Share Europe](https://gs.statcounter.com/os-market-share/all/europe/#monthly-200902-202405)
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r}

library(dplyr)
library(tidyr)
library(ggplot2)
library(strucchange)
library(zoo)
library(lmtest)
library(car)
library(pheatmap)
library(orcutt)
```

### Reading dataset 
```{r}
eurostat_nr_computers_raw <- read.csv("eurostat_computer_numbers.csv")
eurostat_households_raw <- read.csv("eurostat_size_of_houses.csv")

os_combined <- read.csv("os_combined-eu-monthly-200902-202405.csv")
windows_combined <- read.csv("windows_version-eu-monthly-200902-202405.csv")
```

### Processing dataset
Data from Eurostat <br> 
Processing  %of households with the computer in UE
```{r}
nr_eu_computers <- subset(eurostat_nr_computers_raw, 
                          freq.indic_is.unit.hhtyp.geo.TIME_PERIOD == 'A,H_COMP,PC_HH,TOTAL,EU27_2007')

nr_eu_computers_year <- list()
for (col in names(nr_eu_computers)[-1]) {
  nr_eu_computers_year[[sub("X", "", col)]] <- as.numeric(nr_eu_computers[1, col])
}

nr_eu_computers <- data.frame(
  TIME_PERIOD = as.numeric(names(nr_eu_computers_year)), 
  OBS_VALUE_COMP = unlist(nr_eu_computers_year)           
)

```
Filling %of computers missing data 
```{r}
#missing data for 2014 and 2016 - filling with mean to keep the trend 
get_mean_by_year <- function(df, year_previous, year_next) {
  mean_value <- df %>%
    filter(TIME_PERIOD %in% c(year_previous, year_next)) %>%  
    summarise(mean_value = mean(OBS_VALUE_COMP, na.rm = TRUE))
  
  return(mean_value$mean_value)
}

nr_eu_computers <- bind_rows(nr_eu_computers, 
                          data.frame(TIME_PERIOD = 2014, OBS_VALUE_COMP = get_mean_by_year(nr_eu_computers, 2013, 2015)))
nr_eu_computers <- bind_rows(nr_eu_computers, 
                          data.frame(TIME_PERIOD = 2016, OBS_VALUE_COMP = get_mean_by_year(nr_eu_computers, 2015, 2017)))

```
Number of households with the computer in EU 
```{r}
#get nr of households in UE per year 
eurostat_households <- eurostat_households_raw %>%
  filter(agechild == 'TOTAL', 
         n_child == 'TOTAL', 
         hhcomp == 'TOTAL', 
         geo == 'EU27_2020') %>%
  select(TIME_PERIOD, OBS_VALUE)

eurostat_households_with_comp <- inner_join(eurostat_households, nr_eu_computers, by = "TIME_PERIOD")

eurostat_households_with_comp <- eurostat_households_with_comp %>%
  mutate(nr_of_households_with_comp = OBS_VALUE * OBS_VALUE_COMP * 10) # Multiplied by 10 because OBS_VALUE*1000 and OBS_VALUE_COMP in %

```

Data from StatCounter 
```{r}
get_year <- function(df) {
  df %>%
    mutate(Date = as.Date(paste0(Date, "-01"))) %>%
    mutate(year = as.integer(format(Date, "%Y")))
}

os_combined <- get_year(os_combined)
windows_combined <- get_year(windows_combined)

windows_combined <- windows_combined %>%
  filter(year <= 2017, year >= 2014)

os_real_volume_raw <- inner_join(os_combined, eurostat_households_with_comp, by = c("year" = "TIME_PERIOD"))

os_real_volume_raw <- os_real_volume_raw %>%
  mutate(households_with_windows = Windows * nr_of_households_with_comp / 100)

windows_real_volume <- os_real_volume_raw %>%
  select(Date, households_with_windows)

windows_df <- inner_join(windows_combined, windows_real_volume, by = "Date")
```

Number of households with specific Windows OS version 
```{r}
columns_to_multiply <- c("Win7", "WinXP", "Win10", "WinVista", "Win8.1", "Win8", "Win2000")
windows_df <- windows_df %>%
  mutate(across(all_of(columns_to_multiply), ~ . * households_with_windows))
```

```{r echo=T, results='hide'} 
summary(windows_df)
```
```{r echo=T, results='hide'} 
head(windows_df)
```

### Data visualisation
```{r}
df_melted <- windows_df %>%
  select(Date, all_of(columns_to_multiply)) %>%
  pivot_longer(cols = -Date, names_to = "Windows_Version", values_to = "Value")

line_plot <- ggplot(df_melted, aes(x = Date, 
                                   y = Value, 
                                   color = Windows_Version, 
                                   group = Windows_Version)) +
  geom_line() +
  labs(title = "Estimated Number of Households Using Specific Windows OS Version",
       x = "Date",
       y = "Number of Households",
       color = "Windows Version") +
  theme_minimal()

ggsave("nr_of_households.png", line_plot, width = 10, height = 4, units = "in")

print(line_plot)
```

```{r}
df_melted <- windows_combined %>%
  select(Date, all_of(columns_to_multiply)) %>%
  pivot_longer(cols = -Date, names_to = "Windows_Version", values_to = "Value")

bar_plot <- ggplot(df_melted, aes(x = Date, y = Value, fill = Windows_Version)) +
  geom_col(position = "stack", width = 22) +
  labs(title = "Estimated Share of Windows OS Versions in Europe",
       x = "Date",
       y = "Percentage Share",
       fill = "Windows Version") +
  theme_minimal() +
  scale_fill_viridis_d(option = "C", begin = 0.1, end = 0.9) +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

print(bar_plot)
```

### Checking structural breakpoints
Chow test for structural breakpoint after the Launch of Windows 10 
```{r}
WINDOW_10_LAUNCH <- '2015-07-01'

windows_df <- windows_df %>%
  mutate(x = row_number())

point <- windows_df %>%
  filter(Date == WINDOW_10_LAUNCH)

test_statistic <- vector("numeric", length(columns_to_multiply))
p_value <- vector("numeric", length(columns_to_multiply))


for (i in seq_along(columns_to_multiply)) {
  col <- columns_to_multiply[i]
  test <- sctest(windows_df[[col]] ~ windows_df$x, type = "Chow", point = point$x)
  test_statistic[i] <- test$stat
  p_value[i] <- test$p.value
}

chow_test_structural_point <- data.frame(
  Column = columns_to_multiply,
  F_Statistic = test_statistic,
  P_Value = p_value, 
  weighted_test_static = test_statistic * unlist(point[columns_to_multiply][1, ])
)

chow_test_structural_point <- chow_test_structural_point %>%
  filter(Column != 'Win10') %>%
  arrange(desc(weighted_test_static))

print(chow_test_structural_point)
```
### Checking correlations 
```{r}
products_df_before <- windows_df %>% 
  filter((x >= point$x - 11) & (x < point$x + 1)) %>% 
  select(all_of(columns_to_multiply))

products_df_before <- products_df_before[, !names(products_df_before) %in% "Win10"]

products_df_after <- windows_df %>% 
  filter((x > point$x + 1) & (x < point$x + 13)) %>% 
  select(all_of(columns_to_multiply))

data_list <- list(before = products_df_before, after = products_df_after)

for (name in names(data_list)) {
  product_df <- data_list[[name]]
  correlation_matrix <- cor(product_df)
  
  pheatmap(correlation_matrix, 
           display_numbers = TRUE, 
           color = colorRampPalette(c('white', 'red'))(100), 
           cluster_rows = FALSE, 
           cluster_cols = FALSE, 
           fontsize_number = 15,
           main = sprintf("Heatmap of Correlation Matrix %s Launch of Windows 10", name))
}
```


```{r}
for (col in columns_to_multiply){
  df_hist <- windows_df %>% 
    filter((x >= point$x - 11) & (x < point$x + 12)) 
    
  hist(df_hist[[col]], main = sprintf("Histogram of %s", col), xlab = col)
}
```

### Modeling effect using regression with discontinuity 
```{r}
evaluate_model <- function(model) {
  print(summary(model))
  print(vif(model))
  print(bptest(model))
  print(dwtest(model))
  plot(model, which = 1)
}

show_predictions <- function(df, point_break, model) {
  df$fit <- predict(model)
  
  beta_2 <- coef(model)["D"]
  beta_3 <- coef(model)["K"]
  
  print(sprintf("beta2: %s", beta_2))
  print(sprintf("beta3: %s", beta_3))
  
  df$predicted_effect <- ifelse(df$x >= point_break, beta_2 + beta_3 * (df$x - point_break), NA)
  df$x_visual <- df$x - df$x[1] - 11.5
  
  graph <- ggplot(df, aes(x = x_visual, y = y)) +
    geom_point() +
    geom_line(aes(y = ifelse(df$x >= point_break, df$fit, NA)), color = 'gray', size = 1) +
    geom_line(aes(y = ifelse(df$x < point_break, df$fit, NA)), color = 'gray', size = 1) +
    geom_line(aes(y = ifelse(df$x >= point_break, df$fit - df$predicted_effect, NA)), color = 'gray', linetype = "dashed", size = 1) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
    ggtitle(sprintf("Users of %s with RDD", col)) +
    xlab("Months after Win10 launch") +
    ylab(sprintf("Users %s", col)) + 
    theme_minimal()
  
  plot(graph)
  return(df)
}
```

Windows7 <> Windows10 
```{r}
result_list <- list()

col <- "Win7"
point_break <- point$x + 1

df <- data.frame(
  x = windows_df$x, 
  y = windows_df[[col]], 
  t = windows_df$households_with_windows,
  wv = windows_df$WinVista,
  wx = windows_df$WinXP,
  w2k = windows_df$Win2000,
  w7 = windows_df$Win7
)

df <- df[(point_break - 12):(point_break + 12),]

df$D <- ifelse(df$x >= point_break, 1, 0)
df$K <- ifelse(df$x >= point_break, df$x - point_break, 0)
df$rest <- df$wv + df$wx + df$w2k

model <- lm(y ~ D + K + t + rest, data = df)
evaluate_model(model)

#Cochrane-Orcutt transformation
co_model <- cochrane.orcutt(model)
summary(co_model)

df <- show_predictions(df, point_break, co_model)

predictions_w7 <- df$predicted_effect
df$model <- col
result_list[[col]] <- df
```
Windows8.1 <> Windows10 
```{r}
col <- "Win8.1"
predictions_w7[is.na(predictions_w7)] <- 0

point_break <- point$x + 1

df <- data.frame(
  x = windows_df$x, 
  y = windows_df[[col]], 
  t = windows_df$households_with_windows,
  wv = windows_df$WinVista,
  wx = windows_df$WinXP,
  w2k = windows_df$Win2000,
  w7 = windows_df$Win7,
  w8 = windows_df$Win8
)

df <- df[(point_break - 12):(point_break + 12),]

df$D <- ifelse(df$x >= point_break, 1, 0)
df$K <- ifelse(df$x >= point_break, df$x - point_break, 0) 
df$rest <- df$wx + df$wv + df$w2k + df$w7 - predictions_w7 + df$w8

model <- lm(y ~ D + K + t + rest, data = df)
evaluate_model(model)

#Cochrane-Orcutt transformation
co_model <- cochrane.orcutt(model)
summary(co_model)

df <- show_predictions(df, point_break, co_model)
df$model <- col

result_list[[col]] <- df
```


Merging predictions
```{r}
expand_predictions <- function(all_predictions, df, model_name) {
  pred_df <- all_predictions %>%
    filter(model == model_name) %>%
    select(x, predicted_effect) %>%
    mutate(predicted_effect = abs(predicted_effect))
  
  df <- inner_join(df, pred_df, by = "x")
  colnames(df)[colnames(df) == 'predicted_effect'] <- sprintf('predicted_effect_%s', sub('.1', '', model_name))
  
  return(df)
}

all_predictions <- bind_rows(result_list)

windows_all_predictions <- expand_predictions(all_predictions, windows_df, 'Win8.1')
windows_all_predictions <- expand_predictions(all_predictions, windows_all_predictions, 'Win7')

windows_all_predictions <- windows_all_predictions %>%
  mutate(windows_new_users = Win10 - predicted_effect_Win8 - predicted_effect_Win7) %>%
  mutate(windows_new_users = ifelse(windows_new_users < 0, 0, windows_new_users))
```

Showing Predictions 
```{r}
legend_names <- c("New and other versions users", "Users from Win8.1", "Users from Win7")

df_melted <- windows_all_predictions %>%
  select(Date, all_of(c("windows_new_users", "predicted_effect_Win8", "predicted_effect_Win7"))) %>%
  rename(
    !!legend_names[1] := windows_new_users,
    !!legend_names[2] := predicted_effect_Win8,
    !!legend_names[3] := predicted_effect_Win7
  ) %>%
  pivot_longer(cols = -Date, names_to = "Windows_Version", values_to = "Value")

plot <- ggplot(df_melted, aes(x = Date, y = Value, fill = Windows_Version)) +
  geom_col(position = "stack", width = 22) +
  labs(
    title = "Estimated Origin of Windows 10 Users in Europe",
    x = "Date",
    y = "Number of households",
    fill = "Windows Version"
  ) +
  theme_minimal() +
  scale_fill_viridis_d(option = "C", begin = 0.1, end = 0.9) +  
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
plot(plot)
```

```{r}
df_new_users_vs_households_with_windows <- data.frame(
  rolling_difference_total_windows_households = c(0, diff(windows_all_predictions$households_with_windows)), 
  new_users_and_other_windows = windows_all_predictions$windows_new_users
) %>%
  filter(new_users_and_other_windows > 0)

print(df_new_users_vs_households_with_windows)
```
